---
title: "MA4605: Lab 2 Report"
author: "Name & ID Number"
date: "Date of lab session"
output: 
  html_document: 
    highlight: pygments
    theme: spacelab
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
library(tidyverse)
library(kableExtra)
```

* * * 

First read in the data:

```{r}
dat = read.csv(file="ExampleData.csv")
```


## Recoding variables  

The ```Stats.ability``` and ```Stats.interest``` variables are recorded as numbers which represent each category of the Likert scales. We need to recode these variables from numbers to category labels. Let's start with the ```Stats.ability``` variable. The ```recode_factor``` command creates a labelled factor from the corresponding numbers. 

```{r}

dat$Stats.ability = recode_factor(dat$Stats.ability, `1` = "very poor", `2` = "poor", `3` = "good", `4` = "very good", `5` = "excellent")

```  

#### Exercise 1: 

```{r}
# Write your R code for Exercise 1 here.


```

## Summarise and plot categorial variables

### Create a frequency distribution

To create a frequency distribution for the ```Stats.ability``` variable use the following command
```{r}

dat %>%
  count(Stats.ability) %>%
  mutate(prop = prop.table(n), pct = prop.table(n)*100) %>%
  kable() %>% # This line of code makes the table look nicer
  kable_styling() # This line of code makes the table look nicer

```


#### Exercise 2: 
Create a frequency distribution for the ```Stats.interest``` variable and comment.   

```{r}
# Write your R code for Exercise 2 here.


```

[Write your comment for Exercise 2 here.]


### Create a barchart

Create a barchart plotting the frequencies on the y-axis:

```{r, fig.height=4, fig.width=4}
  ggplot(aes(x=Stats.ability), data=dat) + geom_bar(fill="#880011") + xlab("Statistics ability rating")

``` 

Create a barchart with the relative frequencies on the y-axis: 

```{r, fig.height=4, fig.width=4}

dat %>%
  count(Stats.ability) %>%
  mutate(prop = prop.table(n)) %>%
  ggplot(aes(x=Stats.ability, y=prop)) + geom_bar(stat="identity", fill="#880011") + 
  xlab("Statistics ability rating")

```

#### Exercise 3: 
Create a barchart for the ```Stats.interest``` variable. You can choose whether to plot the frequency or relative frequency on the y-axis. Play around with the colour of the bars. 

```{r, fig.height=4, fig.width=4}
# Write your R code for Exercise 3 here.


```


## Summarise and plot numeric variables

### Create summary statistics

To create summary statistics for the variable ```LC.points``` use the following code:

```{r}
dat %>%
  summarise(min = min(LC.points,na.rm=TRUE), max = max(LC.points,na.rm=TRUE),
            mean = mean(LC.points,na.rm=TRUE), med = median(LC.points,na.rm=TRUE),
            sd = sd(LC.points,na.rm=TRUE), 
            q1 = quantile(LC.points, probs = 0.25,na.rm=TRUE),
            q3 = quantile(LC.points, probs = 0.75,na.rm=TRUE), IQR = IQR(LC.points,na.rm=TRUE)) %>%
  kable() %>% # This line of code makes the table look nicer
  kable_styling() # This line of code makes the table look nicer

```

### Create a histogram and boxplot

The two plots that are appropriate for numeric variables are a histogram and boxplot. 

To create a histogram with 7 bins:

```{r, fig.height=4, fig.width=4}

ggplot(aes(x=LC.points), data=dat) + geom_histogram(bins=7)

```

To create a boxplot:

```{r, fig.height=4, fig.width=4}

ggplot(aes(y=LC.points), data=dat) + geom_boxplot()

```

#### Exercise 4: 
Create a summary statistics, a histogram and a boxplot for the ```Num.times.exercise``` variable and comment on the results. 

```{r}
# Write your R code for Exercise 4 here.


```

[Write your comment for Exercise 4 here.]