---
title: "MA4605 Lab 1 Report"
author: "Name & ID Number"
date: "Date of lab session"
output: 
  pdf_document
---

 <!-- DO NOT ALTER THE CODE IN THIS CHUNK -->
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)
library(dplyr)
library(ggplot2)
library(knitr)
# install.packages(kableExtra)
library(kableExtra)
```

* * * 
## Lab report

#### Exercise 1: 

```{r}
# enter your code for Exercise 1 here
## we can use the $ sign with the col name to get the first var
whiteside$Insul
```

We have information about when insulation was put in and the temperate and gass usages before and after.

```{r, echo = TRUE}

```

We can see from the above plot that as temperature decreases the gass usage increases. 

#### Exercise 2:

We can also note that gass usage **dropped** the insulation was installed in the house.


```{r}

```

#### Exercise 3:

```{r}
# enter your code for Exercise 3 here


```


#### Exercise 4:

We can see from the following summary measure that the mean amount of gas used is less after insulation was installed. 

```{r}
# enter your code for Exercise 4 here

```

